---
title: "data manipulation hands-on"
author: "<add your name here>"
date: "<add date here>"
output: word_document
---

```{r libraries}
library(PKPDmisc)
library(readr)
library(dplyr)
library(knitr)
```

## DATA IMPORT


```{r}

```


Objectives:

* Import datasets and documents
* Perform basic data manipulation upon importing the data.

### Task-I

Use the .csv files `demog`, `IV`, and `Oral` provided into the data object folder. 

1. Read in all three csv files  



## DATA MANIPULATION
The goals of this section:

* Use data manipulation tools to prepare the dataset for analysis

### Task-II
1.  Rename "DV" column as "COBS"

```{r}

```


2. Add a Formulation column and label IV/Oral for each dataset

3. Create one integrated dataset with both IV and Oral data.

4. Appropriately merge the demographics dataset into the IV and Oral dataset

5. Perform the following tasks:
    a. Ensure that the following columns are numeric and not text: TIME, COBS, WT, AGE, AMT and DOSEs
    b. Change the following:
    d. Filter the dataset such that you remove all rows that were non-numeric
    e. Create a new column called "GENDER" where:
        i. Female = 0
        ii. Male = 1 
    f. Create a new column called RACEN where:
        i. Caucasian = 0
        ii. Asian = 1
        iii. Black = 2
        iv. Hispanic = 3
    g. Remove the following columns
        i. SEX
        ii. RACE
6. Save the above modifications as a new csv file

## Descriptive Statistics

Objectives

* How to make summaries of the data using descriptive statistics and other data manipulation tools (dplyr, base R functions etc)

### Task III


1. Add a column with the count the number of males/females in the dataset 

```{r}

```


2. Summarize counts stratified by gender and race

```{r}

```

4. summarize the min, mean, and max values for WT, AGE:

    a. non-stratified

    b. by Gender

5. What is the Average numbers samples (observations) per individual in this dataset. Hint: make sure you are *only* counting samples, not necessarily all rows are observations!


6. Calculate the Mean, 5th, and 95th percentile concentration at each time point for each formulation and dose level. hint: you can use `?quantile` to calculate various quantiles

```{r}

```


