---
title: "data manipulation hands-on"
author: "Devin Pastoor"
date: "Sept 26, 2016"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r libraries}
library(PKPDmisc)
library(readr)
library(dplyr)
library(knitr)
library(ggplot2)
```

```{r}
opts_chunk$set(echo = FALSE, warnings = F, message = F, fig.width = 8, fig.height = 6)
```

## DATA IMPORT


Objectives:

* Import datasets and documents
* Perform basic data manipulation upon importing the data.

### Task-I

Use the .csv files `demog`, `IV`, and `Oral` provided into the data object folder. 

1. Read in all three csv files  

```{r}
iv_raw <- read_csv("../data/IV.csv")
oral_raw <- read_csv("../data/ORAL.csv")
demog_raw <- read_csv("../data/demog.csv")
```
```{r}
iv_raw %>% head %>% kable
oral_raw %>% head %>% kable
demog_raw %>% head %>% kable
```




## DATA MANIPULATION
The goals of this section:

* Use data manipulation tools to prepare the dataset for analysis

### Task-II

1. Add a Formulation column and label IV/Oral for each dataset

```{r}
iv <- iv_raw %>% mutate(FORM = "IV")
oral <- oral_raw %>% mutate(FORM = "ORAL")
```


2. Create one integrated dataset with both IV and Oral data.

3.  Rename "DV" column as "COBS"

```{r}
pkdat <- bind_rows(iv, oral) %>% rename(COBS = DV)
```

4. Appropriately merge the demographics dataset into the IV and Oral dataset

Individuals only found in PK:

```{r}
anti_join(pkdat, demog_raw) %>% kable()
```

Individuals found only in demographics:

```{r}
anti_join(demog_raw, pkdat) %>% kable()
```

Given missing PK information, we will only retain individuals that also have PK data

```{r}
pkdemog <- left_join(pkdat, demog_raw)
```
```{r}
pkdemog %>% head %>% kable
```

5. Perform the following tasks:

    a. Ensure that the following columns are numeric and not text: TIME, COBS, WT, AGE, AMT and DOSEs
    b. Change the following:
    c. fix COBS to be numeric
    d. Filter the dataset such that you remove all rows that were non-numeric
    e. Create a new column called "GENDER" where:
        i. Female = 0
        ii. Male = 1 
    f. Create a new column called RACEN where:
        i. Caucasian = 0
        ii. Asian = 1
        iii. Black = 2
        iv. Hispanic = 3
        
```{r}
str(pkdemog)
```

```{r}
pkdemog <- pkdemog %>% 
  mutate(
    NNFLG = ifelse(is.na(COBS), 0, 
                   ifelse(!is.na(as.numeric(COBS)), 0, 1))
  )

pkdemog %>% filter(NNFLG == 1) %>% kable()
```
```{r}
pkdemog_cleaned <- pkdemog %>% 
  filter(NNFLG == 0) %>%
  mutate(COBS = as_numeric(COBS))
```


6. Save the above modifications as a new csv file, dropping the columns SEX and RACE

```{r eval = F}
write_nonmem(pkdemog_cleaned, "../data/pkdemog_cleaned.csv")
```

## Descriptive Statistics

Objectives

* How to make summaries of the data using descriptive statistics and other data manipulation tools (dplyr, base R functions etc)

### Task III


1. Summarize counts stratified by gender and race

```{r}
count(pkdemog_cleaned %>% 
        distinct(ID, .keep_all = T), 
      SEX, RACE) %>% kable()
```

2. Add a column with the count the number of males/females in the dataset 

```{r}
gender_counts <- pkdemog_cleaned %>% 
  distinct(ID, .keep_all=TRUE) %>%
  group_by(SEX) %>%
  summarize(CNTGEN = n())

gender_counts %>% kable()
```


```{r}
pkdemog_cleaned %>% left_join(gender_counts) %>% 
  distinct(ID, .keep_all = T) %>% 
  head %>% kable()
```



4. summarize the min, mean, and max values for WT, AGE:

    a. non-stratified (WT only)
    b. by Gender

```{r}
pkdemog_cleaned %>% 
  summarize(minWT = min(WT), 
            meanWT = mean(WT),
            maxWT = max(WT)) %>% kable()
```

```{r}
pkdemog_cleaned %>% 
  group_by(SEX) %>%
  summarize_at(vars(AGE, WT), funs(min, mean, max)) %>% kable()
```

5. Calculate the Median, 5th, and 95th percentile concentration at each time point for each formulation and dose level. hint: you can use `?quantile` to calculate various quantiles

```{r}
quantile_summaries <- pkdemog_cleaned %>% filter(!is.na(COBS)) %>%
  group_by(FORM, DOSE, TIME) %>%
  summarize(q05 = quantile(COBS, 0.05),
            q50 = quantile(COBS, 0.5),
            q95 = quantile(COBS, 0.95))

quantile_summaries %>% kable()
```


## BONUS: visualize the concentration time data with the quantiles underlaid

```{r}
gg_form_dose <- pkdemog_cleaned %>%
  filter(!is.na(COBS)) %>%
  ggplot(aes(x = TIME, y = COBS, group = interaction(ID, FORM))) +
  geom_ribbon(data = quantile_summaries, 
              aes(ymin = q05, 
                  ymax = q95, 
                  ## since don't have a y and group in this dataset but ggplot
                  ## will look since they are set in the base layer, need to set
                  ## them to null
                  y = NULL, 
                  group = NULL), 
              alpha = 0.9, fill = "blue") + 
  geom_line() +
  geom_point() 

```

```{r}
gg_form_dose +
  facet_wrap(FORM~DOSE)
```

```{r}
gg_form_dose +
  facet_wrap(FORM~DOSE, scales = "free") +
  theme_bw() +
  base_theme() +
  labs(
    x = "Time, hrs",
    y = "Concentration, mg/L"
  )
```

